---
title: "Statistical Learning: Hands-On Workshop"
subtitle: "ISLR §2.1 — Companion Code"
author: "Bolong Tang · Peize Zhang · Yi Zhi Zhang"
date: today
format:
  html:
    toc: true
    code-fold: false
    code-tools: true
execute:
  warning: false
  message: false
---

## Before We Start

This is a ~10-minute hands-on companion to the slides. Work through each
section in order. Every chunk below runs as-is — the **YOUR TURN** callouts
ask you to change a value and re-run the chunk to see what happens.

```{r}
#| label: setup
# install.packages(c("class", "FNN", "ggplot2"))
library(class)   # KNN classification
library(FNN)     # KNN regression
library(ggplot2)
theme_set(theme_minimal())
```

## 1. Estimating *f*: Simple Linear Regression

Recall from the slides: exam score $Y$ as a function of hours studied $X$,

$$ Y = f(X) + \epsilon, \qquad f(X) = \beta_0 + \beta_1 X $$

```{r}
#| label: exam-data
set.seed(2026)
n     <- 60
hours <- runif(n, 0, 10)
score <- 50 + 4.5 * hours + rnorm(n, sd = 8)
exam  <- data.frame(hours, score)

fit_lm <- lm(score ~ hours, data = exam)
summary(fit_lm)
```

```{r}
#| label: exam-plot
ggplot(exam, aes(hours, score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(title = "Simple Linear Regression: Exam Score vs. Hours Studied",
       x = "Hours Studied", y = "Exam Score")
```

```{r}
#| label: exam-predict
predict(fit_lm, newdata = data.frame(hours = 6))
```

> **YOUR TURN:** What are $\hat\beta_0$ and $\hat\beta_1$ from the summary
> above? Change `hours = 6` in the last chunk to a few other values and
> re-run — does the predicted score change the way you'd expect from the
> slope?

## 2. Logistic Regression: Spam Detection

```{r}
#| label: spam-data
set.seed(2026)
n          <- 200
link_count <- rpois(n, lambda = 2)
caps_ratio <- runif(n, 0, 1)
z    <- -2 + 1.1 * link_count + 3 * caps_ratio
p    <- 1 / (1 + exp(-z))
spam <- rbinom(n, 1, p)
email <- data.frame(link_count, caps_ratio, spam)

fit_glm <- glm(spam ~ link_count + caps_ratio, data = email, family = binomial)
summary(fit_glm)
```

```{r}
#| label: spam-predict
new_email <- data.frame(link_count = 4, caps_ratio = 0.6)
predict(fit_glm, newdata = new_email, type = "response")
```

> **YOUR TURN:** Edit `link_count` and `caps_ratio` in `new_email` above.
> Can you find a combination where the predicted probability crosses 0.5
> (the usual classification threshold)?

## 3. KNN Works for Both

Same setup as the house example on the slides: predict a new house's price
(regression) and whether it sold above asking (classification) from its
5 nearest neighbors.

```{r}
#| label: house-data
house <- data.frame(
  sqft  = c(1750, 1820, 1690, 1905, 1760, 1650, 2000, 1580, 1880, 1720),
  age   = c(12, 8, 20, 5, 15, 25, 3, 30, 10, 18),
  price = c(410, 430, 395, 460, 415, 380, 480, 360, 445, 405),
  above_asking = factor(c("Yes","Yes","No","Yes","No","No","Yes","No","Yes","No"))
)

new_house <- data.frame(sqft = 1800, age = 15)
```

```{r}
#| label: knn-regression
# KNN regression: predict price
knn_price <- knn.reg(train = house[, c("sqft", "age")],
                      test  = new_house,
                      y     = house$price,
                      k     = 5)
knn_price$pred
```

```{r}
#| label: knn-classification
# KNN classification: predict above/below asking
knn_class <- knn(train = house[, c("sqft", "age")],
                  test  = new_house,
                  cl    = house$above_asking,
                  k     = 5)
knn_class
```

> **YOUR TURN:** Re-run both chunks with `k = 1` and `k = 9` (just edit the
> `k = ` argument). How does the price prediction move? Also — `sqft` and
> `age` are on very different scales, so distance is dominated by `sqft`.
> Try `scale(house[, c("sqft","age")])` in place of the raw columns and see
> if the prediction changes.

## 4. Bias-Variance Tradeoff, Visually

```{r}
#| label: bias-variance
set.seed(42)
x_train <- sort(runif(40, 0, 10))
y_train <- sin(x_train) + rnorm(40, sd = 0.3)
x_grid  <- seq(0, 10, length.out = 200)

fits <- lapply(c(1, 5, 20), function(k) {
  pred <- knn.reg(train = matrix(x_train), test = matrix(x_grid),
                   y = y_train, k = k)$pred
  data.frame(x = x_grid, y = pred, K = paste0("K = ", k))
})
fits_df <- do.call(rbind, fits)

ggplot() +
  geom_point(data = data.frame(x_train, y_train),
             aes(x_train, y_train), alpha = 0.4) +
  geom_line(data = fits_df, aes(x, y, color = K), linewidth = 1) +
  labs(title = "KNN Regression at Different K", x = "X", y = "Y", color = NULL)
```

> **YOUR TURN:** Which K looks like it's overfitting the training points?
> Which one is too smooth (underfitting)? This is the same tradeoff from the
> slides — small K → low bias / high variance; large K → high bias / low
> variance.

## 5. Wrap-Up Discussion (2 min)

- For the exam-score data, would you trust the linear model or KNN more? Why?
- For the spam classifier, what happens if a new email's features fall far
  outside the training range?
- Go back through sections 1–4: for each example, is it a **regression** or
  **classification** problem, and is the goal **prediction** or
  **inference**?
